\bf Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2024-08-25 paper.qmd
Updated Figure 17

<!--add langevitour screenshots and youtube animation link-->
```{r}
trimesh_removed_mnist 
```

::: {#fig-mnist1-sc layout-ncol="4" fig-pos="H"}
![](figures/mnist/mnist_model_2d.png){#fig-mnist1-model}

![](figures/mnist/sc_1.png){#fig-mnist1-sc1}

![](figures/mnist/sc_2.png){#fig-mnist1-sc2}

![](figures/mnist/sc_3.png){#fig-mnist1-sc3}

Model in \gD{}, on the PaCMAP layout, and three views of the fit in projections from $10\text{-}D$, for the digit 1 of MNIST data ($(s_1, \ s_2) = (-0.100, \ -0.059)$, $b = 374 \  (22, \ 17)$, $m = 140$, benchmark value to remove large edges is $0.094$). (The **langevitour** software is used to view the data with a tour, and the full video is available at <https://youtu.be/zcg_GXBmqjA>).

:::
<!-- need to update the youtube recording-->

>>>>

```{r}
#| label: mnist-model-proj

# Apply the scaling
scaled_mnist_data <- scale_data_manual(training_data_mnist |> select(-ID)) |>
  as_tibble()

df_b_mnist <- df_bin_mnist |>
  dplyr::filter(hb_id %in% df_bin_centroids_mnist$hexID) |>
  dplyr::mutate(type = "model") ## Data with summarized mean

## Reorder the rows of df_b according to the hexID order in df_b_with_center_data
df_b_mnist <- df_b_mnist[match(df_bin_centroids_mnist$hexID, df_b_mnist$hb_id),] |>
  dplyr::select(-hb_id) |>
  select(-type)

# Apply the scaling
scaled_mnist_data_model <- scale_data_manual(df_b_mnist) |>
  as_tibble()

# Combine with the true model for visualization
df <- dplyr::bind_rows(scaled_mnist_data_model |> mutate(type = "model"),
                       scaled_mnist_data |> mutate(type = "data"))

## First projection
projection <- cbind(
    c(0.71503,0.00502,-0.04370,-0.01811,-0.01103,0.09877,0.06199,0.05396,0.07771,-0.01713),
    c(0.00587,0.71879,0.03485,0.08284,-0.07534,-0.05420,0.04315,-0.00547,-0.02995,0.01810))

projected <- as.matrix(scaled_mnist_data) %*% projection

projected_model <- as.matrix(scaled_mnist_data_model) %*% projection

model_df <- dplyr::left_join(
  distance_df_small_edges_mnist |> select(-distance), 
  projected_model_df, 
  by = c("from" = "ID"))

axes_obj <- gen_axes(
  data = scaled_mnist_data, 
  proj_data = projected_df, 
  limits = 1)

mnist_proj_pacmap_model1 <- projected_df |>
  ggplot(
    aes(
      x = proj1, 
      y = proj2)) +
  geom_segment(
    data = model_df, 
    aes(
      x = proj1_from, 
      y = proj2_from, 
      xend = proj1_to, 
      yend = proj2_to), 
    color = "#33a02c") +
  geom_point(
    size = 0.5,
    alpha = 0.2) +
  scale_color_manual(values = c("#000000", "#33a02c")) +
  geom_segment(
    data=axes, 
    aes(x=x1, y=y1, xend=x2, yend=y2), 
    colour="grey70") +
  geom_text(
    data=axes, 
    aes(x=x2, y=y2, label=rownames(axes)), 
    colour="grey50",
    size = 3) +
  geom_path(
    data=circle, 
    aes(x=c1, y=c2), colour="grey70") +
  coord_fixed() +
  interior_annotation("b")

## Second projection
projection <- cbind(
    c(0.13639,-0.18552,-0.38537,-0.03088,-0.04663,0.15013,-0.28351,-0.28865,-0.03379,-0.38009),
    c(-0.46592,-0.40953,-0.02752,0.12815,0.08356,-0.12991,-0.28307,0.11145,0.08031,0.10800))

projected <- as.matrix(scaled_mnist_data) %*% projection

projected_model <- as.matrix(scaled_mnist_data_model) %*% projection

model_df <- dplyr::left_join(
  distance_df_small_edges_mnist |> select(-distance), 
  projected_model_df, 
  by = c("from" = "ID"))

axes_obj <- gen_axes(
  data = scaled_mnist_data, 
  proj_data = projected_df, 
  limits = 1)

mnist_proj_pacmap_model2 <- projected_df |>
  ggplot(
    aes(
      x = proj1, 
      y = proj2)) +
  geom_segment(
    data = model_df, 
    aes(
      x = proj1_from, 
      y = proj2_from, 
      xend = proj1_to, 
      yend = proj2_to), 
    color = "#33a02c") +
  geom_point(
    size = 0.5,
    alpha = 0.1) +
  scale_color_manual(values = c("#000000", "#33a02c")) +
  geom_segment(
    data=axes, 
    aes(x=x1, y=y1, xend=x2, yend=y2), 
    colour="grey70") +
  geom_text(
    data=axes, 
    aes(x=x2, y=y2, label=rownames(axes)), 
    colour="grey50",
    size = 3) +
  geom_path(
    data=circle, 
    aes(x=c1, y=c2), colour="grey70") +
  coord_fixed() +
  interior_annotation("c")

## Third projection
projection <- cbind(
    c(0.47163,0.21622,0.04027,0.17328,0.03603,0.06488,-0.22226,-0.07186,0.39674,0.13485),
    c(-0.41187,0.36322,0.11264,-0.14175,0.19810,0.07062,0.00680,0.08795,0.37068,-0.11280))

projected <- as.matrix(scaled_mnist_data) %*% projection

projected_model <- as.matrix(scaled_mnist_data_model) %*% projection

model_df <- dplyr::left_join(
  distance_df_small_edges_mnist |> select(-distance), 
  projected_model_df, 
  by = c("from" = "ID"))

axes_obj <- gen_axes(
  data = scaled_mnist_data, 
  proj_data = projected_df, 
  limits = 1)

mnist_proj_pacmap_model3 <- projected_df |>
  ggplot(
    aes(
      x = proj1, 
      y = proj2)) +
  geom_segment(
    data = model_df, 
    aes(
      x = proj1_from, 
      y = proj2_from, 
      xend = proj1_to, 
      yend = proj2_to), 
    color = "#33a02c") +
  geom_point(
    size = 0.5,
    alpha = 0.2) +
  scale_color_manual(values = c("#000000", "#33a02c")) +
  geom_segment(
    data=axes, 
    aes(x=x1, y=y1, xend=x2, yend=y2), 
    colour="grey70") +
  geom_text(
    data=axes, 
    aes(x=x2, y=y2, label=rownames(axes)), 
    colour="grey50",
    size = 3) +
  geom_path(
    data=circle, 
    aes(x=c1, y=c2), colour="grey70") +
  coord_fixed() +
  interior_annotation("d")

<!--add langevitour screenshots and youtube animation link-->
```{r}
#| label: fig-mnist-tri-proj
#| fig-pos: H
#| fig-height: 10
#| fig-width: 10
#| fig-cap: "Model in \\gD{}, on the PaCMAP layout, and three views of the fit in projections from $10\\text{-}D$, for the digit 1 of MNIST data ($(s_1, \ s_2) = (-0.100, \ -0.059)$, $b = 374 \  (22, \ 17)$, $m = 140$, benchmark value to remove large edges is $0.094$). (The **langevitour** software is used to view the data with a tour, and the full video is available at <https://youtu.be/zcg_GXBmqjA>)."

trimesh_removed_mnist + mnist_proj_pacmap_model1 +
  mnist_proj_pacmap_model2 + mnist_proj_pacmap_model3 +
  plot_layout(ncol = 2)
```

<!-- ::: {#fig-mnist1-sc layout-ncol="4" fig-pos="H"} -->
<!-- ![](figures/mnist/mnist_model_2d.png){#fig-mnist1-model} -->

<!-- ![](figures/mnist/sc_1.png){#fig-mnist1-sc1} -->

<!-- ![](figures/mnist/sc_2.png){#fig-mnist1-sc2} -->

<!-- ![](figures/mnist/sc_3.png){#fig-mnist1-sc3} -->

<!-- Model in \gD{}, on the PaCMAP layout, and three views of the fit in projections from $10\text{-}D$, for the digit 1 of MNIST data ($(s_1, \ s_2) = (-0.100, \ -0.059)$, $b = 374 \  (22, \ 17)$, $m = 140$, benchmark value to remove large edges is $0.094$). (The **langevitour** software is used to view the data with a tour, and the full video is available at <https://youtu.be/zcg_GXBmqjA>). -->

<!-- ::: -->
<!-- need to update the youtube recording-->

### 2024-08-25 paper.tex
Updated Figure 17

According to Figure~\ref{fig-mnist1-sc1}, the non-linear continuous
structure observed in the \gD{} representation of PaCMAP
(Figure~\ref{fig-pacmap-author}) is also visible when visualizing the
model overlaid on the data space. This indicates that PaCMAP accurately
captures the structure of the \pD{} data. Additionally, the model shows
a twisted pattern within the non-linear structure in \pD{} space
(Figure~\ref{fig-mnist1-sc2}), which is an additional pattern not
visible in the \gD{} representation (Figure~\ref{fig-pacmap-author}).
Furthermore, as shown in Figure~\ref{fig-mnist1-sc3}, some long edges
exist in the \pD{} space that are not recognized as long edges in the
\gD{} representation. However, PaCMAP is a reasonable \gD{}
representation of MNIST digit 1 data, because PaCMAP preserves the
non-linear structure present in the \pD{} data.

\begin{center}
\includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/unnamed-chunk-60-1.pdf}
\end{center}

@@

\begin{minipage}{0.25\linewidth}

@@

\includegraphics{figures/mnist/mnist_model_2d.png}

@@

\subcaption{\label{fig-mnist1-model}}

\end{minipage}%
%
\begin{minipage}{0.25\linewidth}

\includegraphics{figures/mnist/sc_1.png}

\subcaption{\label{fig-mnist1-sc1}}

\end{minipage}%
%
\begin{minipage}{0.25\linewidth}

\includegraphics{figures/mnist/sc_2.png}

\subcaption{\label{fig-mnist1-sc2}}

\end{minipage}%
%
\begin{minipage}{0.25\linewidth}

\includegraphics{figures/mnist/sc_3.png}

\subcaption{\label{fig-mnist1-sc3}}

\caption{\label{fig-mnist1-sc}Model in \gD{}, on the PaCMAP layout, and
three views of the fit in projections from \(10\text{-}D\), for the
digit 1 of MNIST data (\((s_1, \ s_2) = (-0.100, \ -0.059)\),
\(b = 374 \  (22, \ 17)\), \(m = 140\), benchmark value to remove large
edges is \(0.094\)). (The \textbf{langevitour} software is used to view
the data with a tour, and the full video is available at
\url{https://youtu.be/zcg_GXBmqjA}).}

>>>>

According to \textbf{?@fig-mnist1-sc1}, the non-linear continuous
structure observed in the \gD{} representation of PaCMAP
(Figure~\ref{fig-pacmap-author}) is also visible when visualizing the
model overlaid on the data space. This indicates that PaCMAP accurately
captures the structure of the \pD{} data. Additionally, the model shows
a twisted pattern within the non-linear structure in \pD{} space
(\textbf{?@fig-mnist1-sc2}), which is an additional pattern not visible
in the \gD{} representation (Figure~\ref{fig-pacmap-author}).
Furthermore, as shown in \textbf{?@fig-mnist1-sc3}, some long edges
exist in the \pD{} space that are not recognized as long edges in the
\gD{} representation. However, PaCMAP is a reasonable \gD{}
representation of MNIST digit 1 data, because PaCMAP preserves the
non-linear structure present in the \pD{} data.

@@

@@

\includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-mnist-tri-proj-1.pdf}

@@

\caption{\label{fig-mnist-tri-proj}Model in \gD{}, on the PaCMAP layout,
and three views of the fit in projections from \(10\text{-}D\), for the
digit 1 of MNIST data (\((s_1,  s_2) = (-0.100,  -0.059)\),
\(b = 374   (22,  17)\), \(m = 140\), benchmark value to remove large
edges is \(0.094\)). (The \textbf{langevitour} software is used to view
the data with a tour, and the full video is available at
\url{https://youtu.be/zcg_GXBmqjA}).}
